---
title: "Urban Ministries Project 3"
author: "Amber Young"
date: "11/18/2019"
output: html_document
---
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
df <- read_csv("../data/cleaned.csv", col_names=T)
df <- df[,2:41]
```
# Introduction
Urban Ministries, established in 1983, offers many programs to help support homeless and low-income people in the Durham community. These programs include:

* A community shelter, which can accomodate over 140 people a night
* A community cafe, which offers breakfast, lunch, and dinner daily to the homeless
* A food pantry, that provides people with groceries, and
* A clothing closet, that supplies individuals and families with clothing each month.

All together this organization helps about 6,000 people per year with over 4,000 volunteers, and they rely heavily on donations of food and clothing from the community.

This data, provided by Urban Ministries, focuses on their emergency shelter program.

# Purpose

The purpose of this project is to provide UM with summary statistics and plots describing the characteristics of people who have used their emergency shelter. Additionally, models are built to predict whether someone will complete the program, and how long they will stay.


# Characteristics of Urban Ministries Shelter Clients
This section provides some generally information about the characteristics of UM Clients.

## Age
In the plot below, we have the number of UM Clients by age at entry to the shelter. We can see that almost all are under 60 years old and the majority are between 40 and 60. There is also a large amount of people in their 20s.
```{r echo=FALSE, message=FALSE, warning=F}
ggplot(df, aes(x=Age_Entry))+
  geom_histogram(bins=20, fill="darkturquoise", color="black")+
  labs(title="Emergency Shelter Clients by Age", x="Age at Entry to Shelter", y="Number of Clients")
```

## Gender
We can see below that the UM shelter serves a much larger proportion of Males than Females, with the male clientele representing over 70%. We can also see that they serve a few from the transgender community.
```{r echo=FALSE, message=FALSE, warning=F}
df3 <- df %>% group_by(Gender) %>% summarise(percent=n()/5299*100)
ggplot(df3, aes(x=reorder(Gender, -percent), y=percent, fill=Gender))+
  geom_bar(width=1, stat="identity", color="black")+
  coord_flip()+
  theme(legend.position="none")+
  labs(title="Percent of Clients by Gender", x="Gender", y="Percent of Clients")
```

## Race
We can see below that there is a much large percentage of Black/African American clients at UM shelters than White clients. There are also a few American Indian/Alaskan Natives who have used the shelter.
```{r echo=FALSE, message=FALSE, warning=F}
df2 <- df[df$Race=="White (HUD)" | df$Race=="Black or African American (HUD)" | df$Race=="American Indian or Alaska Native (HUD)" & !is.na(df$Race),]
df3 <- df2 %>% group_by(Race) %>% summarise(percent=n()/5299*100)
ggplot(df3, aes(x=reorder(Race, -percent), y=percent, fill=Race))+
  geom_bar(width=1, stat="identity", color="black")+
  coord_flip()+
  theme(legend.position="none")+
  labs(title="Percent of Clients by Race", x="Race", y="Percent of Clients")
```

## Disability Status
This plot shows the percentage of clients who do or do not have some form of disability. We can see that there are more disabled people using the shelter than not disabled people.

```{r echo=FALSE, message=FALSE, warning=F}
df_dis <- df[df$Disabling_Condition=='Yes (HUD)' | df$Disabling_Condition=='No (HUD)' | is.na(df$Disabling_Condition),]
df3 <- df_dis %>% group_by(Disabling_Condition) %>% summarise(percent=n()/5299*100)
ggplot(df3, aes(x=reorder(Disabling_Condition, -percent), y=percent, fill=Disabling_Condition))+
  geom_bar(width=1, stat="identity", color="black")+
  coord_flip()+
  theme(legend.position="none")+
  labs(title="Percent of Clients by Disability Status", x="Disability Status", y="Percent of Clients")
```

## Reason for Leaving the Shelter
This plot shows the percent of clients who leave the shelter by their reason for leaving. We can see that a large number of them just disappear (over 20%), 10% of them complete the UM program, about 8% leave for a housing opportunity, and the rest leave for a variety of other reasons.
```{r echo=FALSE, message=FALSE, warning=F}
df3 <- df %>% group_by(Reason_for_Leaving) %>% summarise(percent=n()/5299*100)
ggplot(df3, aes(x=reorder(Reason_for_Leaving, -percent), y=percent, fill=Reason_for_Leaving))+
  geom_bar(width=1, stat="identity", color="black")+
  coord_flip()+
  theme(legend.position="none")+
  labs(title="Percent of Clients by Reason for Leaving", x="Reason for Leaving", y="Percent of Clients")
```

## Length of Stay at Shelter
The plot below shows how long people tend to stay at the shelter. We can see that over 65% stay for less than a month, and over 90% stay for less than 100 days. In fact, the average time that a client stays is just about 40 days. The most common occurence is for a client to stay only one day.
```{r echo=FALSE, message=FALSE, warning=F}
df2 <- df[df$days_at_shelter != 0,]
summary(df2$days_at_shelter)
per90 <- quantile(df2$days_at_shelter, probs=.9, na.rm=T)
ggplot(df2, aes(x=days_at_shelter))+
  geom_histogram()+
  geom_vline(xintercept = per90, color="red")+
  labs(title="Number of Clients by How Long They Stay", x="Number of Nights Spent at the Shelter", y="Number of Clients")+
  geom_text(aes(per90-50,2700,label = '90%'))+
  annotate("rect", xmin = -1, xmax = per90, ymin = -1, ymax = 2750,
           alpha = .2, fill="red")
```

# Models and Predictions
More to come...

# Conclusions
Urban Ministries serves a diverse clientele, but some characteristics are more common than others. Knowing this can help them better prepare their facilities for the specific type of clients they see. For example, the majority of their clients are between 40 and 60 years old, and over half of them are disabled in some way. Therefore, job resources they might offer need to target to this slightly older age group, and to those that may be limited either physically or mentally. Additionally, the overwhelming majority of clients are black/African American. This information is very powerful in bringing to light the socioeconomic disparity between races in the Durham area. 

Secondly, we find that a large number of people are disappearing from the shelter without finding housing or completing the UM program, and we see that most people don't stay at the shelter very long.

In the second part of this report, models were built to predict program success and duration of stay based on client characteristics. ...... more to come.